---
title: "Sentiment Analysis of Primeminister Speeches"
authors: Ida, Sofia, Sidse
output: html_document
date: "2026-05-26"
---
Introduction: In this analysis we are examining the emotional value of danish prime ministers speeches. There is a speech from each term of office from the year 1999 to 2025. Only the relevant paragraph about immigrants is taken from each speech.

##Step 1
First install the needed packages. Danish AFINN dictionary is loaded directly from its source repository using a UTF-8 encoding configuration to preserve Danish special characters (æ,ø,å).

```{r, message=FALSE, warning=FALSE}
library(pdftools)
library(tidytext)
library(tidyverse)
library(ggwordcloud)

#the Danish AFINN dataset
url <- "https://raw.githubusercontent.com/fnielsen/afinn/master/afinn/data/AFINN-da-32.txt"

danish_lexicon <- read.delim(url, header = FALSE, 
  col.names = c("word", "value"), 
  stringsAsFactors = FALSE,
  encoding = "UTF-8")%>%
  mutate(word = tolower(word))
```

##Step 2
Have R find the needed pdf file, extract the needed words and make into a clean dataframe of words. 
Using the function unnest_tokens() turns the text into lowercase single-word rows.
By adding the filter(^[a-zæøå]+$) we ensure that it goes by the danish alphabet.


Helle Thorning Schmidt speech from 2012
```{r}
file_to_read <- "data/Thorning_2012.pdf"
raw_text <- pdftools::pdf_text(file_to_read)
speech_dataframe <- data.frame(text = raw_text, stringsAsFactors = FALSE)
cleaned_words <- speech_dataframe %>%
  unnest_tokens(word, text) %>%
  mutate(word = str_remove_all(word, "[[:punct:]]")) %>%
  filter(str_detect(word, "^[a-zæøå]+$"))
```

##Step 3
Here we isolate the emotional words that are present in both the text and the dictionary. With the print() we can see which words it found.

```{r}
matched_words <- cleaned_words %>%
  inner_join(danish_lexicon, by = "word")

print(matched_words %>% select(word, value) %>% distinct())
```
##Step 4
Finally we can calculate the emotional metrics for the speech.  

```{r}
Thorning_summary <- matched_words %>%
  summarize(File_Analyzed = basename(file_to_read),
    Average_Sentiment  = mean(value),
    Total_Scored_Words = n())
print(Thorning_summary)
```
##Step 5
Anders Fogh Rasmussen speech from 2005
Finding the pdf file and cleaning the data
```{r}
file_to_read <- "data/Fogh_2005.pdf"
raw_text <- pdftools::pdf_text(file_to_read)
speech_dataframe <- data.frame(text = raw_text, stringsAsFactors = FALSE)

cleaned_words <- speech_dataframe %>%
  unnest_tokens(word, text) %>%
  mutate(word = str_remove_all(word, "[[:punct:]]")) %>%
  filter(str_detect(word, "^[a-zæøå]+$"))
```

##Step 6
Isolating the emotional words
```{r}
matched_words <- cleaned_words %>%
  inner_join(danish_lexicon, by = "word")

print(matched_words %>% select(word, value) %>% distinct())
```
##Step 7
Calculating the sentiment of the words

```{r}
Fogh_summary <- matched_words %>%
  summarize(File_Analyzed = basename(file_to_read),
    Average_Sentiment  = mean(value),
    Total_Scored_Words = n())
print(Fogh_summary)
```
##Step 8
Lars Løkke Rasmussen speech from 2010
Finding the pdf file and cleaning the data
```{r}
file_to_read <- "data/Løkke_2010.pdf"
raw_text <- pdftools::pdf_text(file_to_read)
speech_dataframe <- data.frame(text = raw_text, stringsAsFactors = FALSE)
cleaned_words <- speech_dataframe %>%
  unnest_tokens(word, text) %>%
  mutate(word = str_remove_all(word, "[[:punct:]]")) %>%
  filter(str_detect(word, "^[a-zæøå]+$"))
```

##Step 9
Isolating the emotional words
```{r}
matched_words <- cleaned_words %>%
  inner_join(danish_lexicon, by = "word")

print(matched_words %>% select(word, value) %>% distinct())
```
##Step 10
Calculating the sentiment of the words
```{r}
Løkke10_summary <- matched_words %>%
  summarize(File_Analyzed = basename(file_to_read),
    Average_Sentiment  = mean(value),
    Total_Scored_Words = n())
print(Løkke10_summary)
```
##Step 11
Lars Løkke Rasmussens speech from 2017
Finding the pdf file and cleaning the data
```{r}
file_to_read <- "data/Løkke_2017.pdf"
raw_text <- pdftools::pdf_text(file_to_read)
speech_dataframe <- data.frame(text = raw_text, stringsAsFactors = FALSE)
cleaned_words <- speech_dataframe %>%
  unnest_tokens(word, text) %>%
  mutate(word = str_remove_all(word, "[[:punct:]]")) %>%
  filter(str_detect(word, "^[a-zæøå]+$"))
```

##Step 12
Isolating the emotional words
```{r}
matched_words <- cleaned_words %>%
  inner_join(danish_lexicon, by = "word")

print(matched_words %>% select(word, value) %>% distinct())
```
##Step 13
Calculating the sentiment of the words
```{r}
Løkke17_summary <- matched_words %>%
  summarize(File_Analyzed = basename(file_to_read),
    Average_Sentiment  = mean(value),
    Total_Scored_Words = n())
print(Løkke17_summary)
```
##Step 14
Mette Frederiksen speech from 2022
Finding the pdf file and cleaning the data
```{r}
file_to_read <- "data/Frederiksen_2022.pdf"
raw_text <- pdftools::pdf_text(file_to_read)
speech_dataframe <- data.frame(text = raw_text, stringsAsFactors = FALSE)
cleaned_words <- speech_dataframe %>%
  unnest_tokens(word, text) %>%
  mutate(word = str_remove_all(word, "[[:punct:]]")) %>%
  filter(str_detect(word, "^[a-zæøå]+$"))
```

##Step 15
Isolating the emotional words
```{r}
matched_words <- cleaned_words %>%
  inner_join(danish_lexicon, by = "word")

print(matched_words %>% select(word, value) %>% distinct())
```
##Step 16
Calculating the sentiment of the words
```{r}
Frederiksen_summary <- matched_words %>%
  summarize(File_Analyzed = basename(file_to_read),
    Average_Sentiment  = mean(value),
    Total_Scored_Words = n())
print(Frederiksen_summary)
```
##Step 17
Poul Nyrup Rasmussen speech from 1999
Finding the pdf file and cleaning the data
```{r}
file_to_read <- "data/Nyrup_1999.pdf"
raw_text <- pdftools::pdf_text(file_to_read)
speech_dataframe <- data.frame(text = raw_text, stringsAsFactors = FALSE)
cleaned_words <- speech_dataframe %>%
  unnest_tokens(word, text) %>%
  mutate(word = str_remove_all(word, "[[:punct:]]")) %>%
  filter(str_detect(word, "^[a-zæøå]+$"))
```

##Step 18
Isolating the emotional words
```{r}
matched_words <- cleaned_words %>%
  inner_join(danish_lexicon, by = "word")

print(matched_words %>% select(word, value) %>% distinct())
```
##Step 19
Calculating the sentiment of the words
```{r}
Nyrup_summary <- matched_words %>%
  summarize(File_Analyzed = basename(file_to_read),
    Average_Sentiment  = mean(value),
    Total_Scored_Words = n())
print(Nyrup_summary)
```
##Step 20
Here we make a graph showing each speech in comparison

```{r final-comparison, message=FALSE, warning=FALSE}
timeline <- bind_rows(
  Thorning_summary, 
  Fogh_summary, 
  Løkke10_summary,
  Løkke17_summary,
  Frederiksen_summary,
  Nyrup_summary) %>%
  mutate(Year = c(2012, 2005, 2010, 2017, 2022, 1999), 
         Prime_Minister = c("Thorning", "Fogh", "Løkke I", "Løkke II", "Frederiksen", "Nyrup")) %>%
  select(Year, Prime_Minister, Average_Sentiment, Total_Scored_Words, File_Analyzed)

knitr::kable(timeline, caption = "Historical Overview of Speech Sentiment")

ggplot(timeline, aes(x = Year, y = Average_Sentiment)) +
  geom_hline(yintercept = 0, color = "grey", linetype = "dashed", linewidth = 0.8) + 
  geom_line(color = "lightblue", linewidth = 1.2) +
  geom_point(color = "orange", size = 4) +
  geom_text(aes(label = Prime_Minister), vjust = -1.5, size = 3.5, fontface = "bold") +
  scale_x_continuous(breaks = seq(2000, 2026, by = 2)) +
  expand_limits(y = c(min(timeline$Average_Sentiment) - 0.5, 
                       max(timeline$Average_Sentiment) + 0.5)) +
  labs(title = "Primeminister speeches (1999-2025)",
    x = "Year of Speech",
    y = "Average Sentiment Score") +
  theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(color = "dimgrey", size = 10, hjust = 0.5),
    panel.grid.minor = element_blank())
```
